LUCID: Learning-Enabled Uncertainty-Aware Certification of Stochastic Dynamical Systems

### 2025-07-30 docs/paper/main.tex
feat: update paper

\documentclass{article}

@@

@@

\title{Delta-Complete Linear Programming}

\author{Oliver Sch\"{o}n \and
Ernesto Casablanca \and
Sadegh Soudjani \and
Paolo Zuliani
}

\begin{comment}
\institute{O. Sh\"{o}n,E. Casablanca \at Newcastle University, Newcastle upon Tyne, United Kingdgom \\ \email{e.casablanca2@newcastle.ac.uk}, \email{o.schoen2@newcastle.ac.uk}
\and
S. Soudjani \at Max Planck Institute for Software Systems, Kaiserslautern, Germany \\ \email{sadegh@mpi-sws.org}
\and
P. Zuliani \at Universit\`{a} di Roma ``La Sapienza'', Rome, Italy\\ \email{zuliani@di.uniroma1.it}
}
\end{comment}

@@

\begin{abstract}
  Ensuring the safety of AI-enabled systems, particularly in high-stakes domains like autonomous driving and healthcare, has become increasingly critical.
  Traditional formal verification tools fall short when faced with the opaque, black-box nature of AI components and the sheer scale of modern applications.
  To address these challenges, we introduce Lucid, a new tool designed specifically for the safety verification and synthesis of black-box systems governed by discrete-time stochastic dynamics.

  Lucid employs a data-driven methodology rooted in control barrier certificates, which are learned directly from system trajectory data to ensure formal safety guarantees.
  We use conditional mean embeddings to embed data into a reproducing kernel Hilbert space (RKHS), where it constructs an RKHS ambiguity set that can be inflated to robustify the result to out-of-distribution behavior.
  To generalize beyond safety, we provide theoretical foundations for applying Lucid to broader classes of temporal logic specifications.

  A key innovation within Lucid is its use of a finite Fourier expansion to reformulate a semi-infinite non-linear optimization problem into a tractable linear program.
  The resulting spectral barrier allows us to leverage the fast Fourier transform to generate the relaxed problem efficiently, offering a scalable yet distributionally robust framework for verifying safety.
  Lucid thus offers a robust and efficient verification framework, able to handle the complexities of modern black-box systems while providing formal guarantees of safety.
\end{abstract}

\section{Introduction}
\label{sec:intro}

Lucid is in implemented in C++ and it is meant to be used as a library.
It provides a set of interfaces that allow users to have an high level of control and flexibility over the verification process.
We also provide a Python wrapper, called PyLucid, to facilitate the integration of the tool into existing workflows and effortlessly leverage well-established libraries such as NumPy~\cite{numpy} and SciPy~\cite{scipy}.
Lucid is available as an open-source project on GitHub at \url{https://github.com/TendTo/lucid}.
More technical details about the implementation can be found in the documentation at \url{https://tendto.github.io/lucid/}.

\section{Problem Statement}
\label{sec:problem-statement}

\subsection{Notation}
\label{sec:notation}

% Spaces and numbers
We denote the sets of positive integers and non-negative reals as $\N_{>0}$ and $\R_{\geq 0}$, respectively.
% Probability theory
Consider a Polish sample space $\X$ \cite{bogachev2007measure}.
Let $(\X,\borel{\X},\P)$ be the underlying probability space equipped with a Borel $\sigma$-algebra $\borel{\X}$ defined over $\X$, i.e., the smallest $\sigma$-algebra containing open subsets of $\X$, and a probability measure $\P$.
For a random variable $X$, let $p_X$ be the pushforward probability measure of $\P$ under $X$ such that $X\sim p_X(\cdotx)$.

The expected value of a function $f(X)$ on $\X$ is written as $\E_{p_X}[f(X)]$. If it is clear from the context, we abbreviate and write $\E[f(X)]$.
We denote the set of all probability measures for a given measurable space $(\X,\borel{\X})$ as $\mathcal{P}(\X)$.

The $n$-dimensional Gaussian measure with mean $\mu\in\mathbb{R}^n$ and covariance matrix $\Sigma\in\mathbb{R}^{n\times n}$ is given by
\begin{equation}
  \label{eq:gaussian-kernel}
  \mathcal N(dx|\mu, \Sigma) := \frac{dx}{ \sqrt{(2\pi)^n \left|\Sigma\right| }}\exp\left[-\frac{1}{2}(x-\mu)\T\Sigma^{-1\!}(x-\mu)\right],
\end{equation}
where $|\Sigma|$ is the determinant of $\Sigma$.
The Dirac delta measure $\delta_a:\borel{\X}\rightarrow [0,1]$ concentrated at a point $a\in\X$ is defined as $\delta_a(A)=1$ if $a\in A$ and $\delta_a(A)=0$ otherwise, for any measurable set $A\in\borel{\X}$.
We denote the uniform distribution over $\X$ as $\mathcal{U}_\X$ with realizations $x\sim\mathcal{U}_\X(\cdotx)$.

For two measurable spaces $(\X,\borel{\X})$ and $(\Y,\borel{\Y})$, a \emph{probability kernel} is a mapping $\p: \X \times \borel{\Y}\rightarrow  [0,1]$ such that $\p(X=x,\cdotx):\borel{\Y}\rightarrow[0,1]$ is a probability measure for all $x\in\X$, and $\p(\cdotx, B): \X\rightarrow [0,1]$ is measurable for all  $B\in\borel{\Y}$.
A probability kernel associates to each point $x\in\X$ a measure denoted by $\p(\cdotx|X=x)$.

% Linear algebra
The transpose of a vector or matrix $A$ is indicated as $A\T$.
Let the $N\times N$ dimensional identity matrix be given by $I_N$.
Let $X_N:=[x_i]_{i=1}^N$ be a column vector with $x_i\in\X$.
We denote the element-wise evaluation of a function $f:\X\rightarrow\R$ on $X_N$ as $f(X_N):=[f(x_i)]_{i=1}^N$.
Similarly, we may write $A = [a_{ij}]_{i,j=1}^N$ to denote a matrix with its elements.

\subsection{Discrete-Time Stochastic Systems}
In this work, we consider systems expressible as Markov decision processes over continuous state and input spaces, formally defined as follows.
\begin{definition}[Markov decision process (MDP)]
  An MDP is a tuple $\M=(\X,\X_0,\U,\Tr)$, comprising a state space $\X\subset\R^n$ with states $x\in\X$; initial states $x_0\in \X_0\subset\X$; an input space $\U$ with inputs $u\in\U$; and a probability kernel $\Tr:\X\times\U\times\borel{\X}\rightarrow[0,1]$.
\end{definition}

In every execution, given a current state $x\in\X$ and control input $u\in\U$, the MDP evolves to a consecutive state $x_{+}\in\X$, which is obtained as a realization $x_{+}\sim\Tr(\,\cdot\,|X=x,U=u)$.
As a notable class of systems that can be captured using MDPs, we may consider black-box systems with Markovian discrete-time stochastic dynamics, namely
\begin{equation}
  \label{eq:model}
  \M: \left\{ \begin{array}{ll}
    x_{t+1} & = f(x_t,u_t,w_t),\quad w_t\sim p_w(\cdotx), \\
    y_t     & = x_t,
  \end{array} \right.
\end{equation}
where the system state, control input, and observation at the $t^{\text{th}}$ time-step are denoted by $x_t$, $u_t$, and $y_t$, respectively.
The state evolution of the system, described by the function $f:\X\times\U\times\mathbb{W}\rightarrow\X$,
is subject to {independent, identically distributed (i.i.d.) noise} $w_t\sim p_w(\cdotx)$ supported on a set $\mathbb{W}$.

\paragraph{Implementation}

When using \lucid, the user needs to provide a set of sampled trajectories from the system.
This is often achieved by defining a function that takes as input the current state and control input and returns the next state, adding an i.i.d. noise sampled from an appropriate distribution, or it may be collected from a simulator or a real-world system.
The sampled trajectories are then fed into \lucid, either programmatically or through a file \todo{CSV, HDF5, other?}.
\todo{Allow a simple way to define the system model? Maybe latex2sympy?}
\todo{Graphical interface?}

\subsection{System Safety}
In this paper, we wish to synthesize a controller $\pi$ such that the closed-loop system $\pi\times\M$ is safe.
Particularly, we restrict our attention to \emph{stationary} policies of the form $\pi:\X\rightarrow\U$, as for MDPs there always exists a stationary policy that is optimal.
For an MDP $\M$, we characterize \emph{safety} as a property of its state. More precisely, a system $\pi\times\M$ is said to be safe if for any initial state $x_0\in \X_0$ the system does not enter an \emph{unsafe set} of states $\X_u\subset\X$.
Note that general stochastic systems do not admit a binary notion of safety.
Thus, we quantify the \emph{probability} of reaching $\X_u$ in a given finite horizon $T\in\N_{>0}$ and compute its complement.
A (finite-horizon) \emph{safety specification} $\psi_{\text{safe}:=(\X_u,T)}$ is hence fully characterized by an unsafe set $\X_u$ and a time horizon $T$.
The system $\pi\times\M$ satisfies $\psi_{\text{safe}}$ with probability at least $p\in(0,1)$ if the probability of its trajectories starting in $\X_0$ and avoiding $\X_u$ within horizon $T$ is at least $p$. This is denoted by $P^\pi_{\text{safe}}(\M):=\P(\pi\times\M\satisfies\psi_{\text{safe}})\geq p$~\cite{BK08}.

\paragraph{Implementation}

\lucid understands \texttt{RectSet}s and \texttt{Multiset}s \todo{PolytopeSet?}.
\todo{Set definition from file?}.
These are used to indicate the initial $\X_0$ set and unsafe set $\X_u$ when defining the safety specification $\psi_{\text{safe}}$.
Once the system is fully defined, the time horizon $T$ is set as a parameter of the verification process.
\todo{Support other kinds of properties other than safety (e.g., reachability)?}

\subsection{Control Barrier Certificates}
\label{sec:cbc}
Whilst certifying the safety of a stochastic system is generally challenging, \emph{control barrier certificates} (CBCs) and \emph{control barrier functions} (CBFs) leverage the concept of set invariance to arrive at an abstraction-free formulation. This has made CBCs/CBFs popular tools for safety verification and synthesis~\cite{prajna2006barrier,cosner2023generative}.
We briefly recall the definition of a CBC.
\begin{definition}[Control barrier certificate (CBC)]\label{def:cbc}
  A function $\B:\X\rightarrow\R_{\geq 0}$ is called a CBC of an MDP $\M=(\X,\X_0,\U,\Tr)$ with reference to an unsafe set $\X_u$, if we have
  \begin{itemize}
    \item[(a)] $\forall x_0\in \X_0\!:\,\B(x_0)\leq\eta$;
    \item[(b)] $\forall x_u\in \X_u\!:\,\B(x_u)\geq\gamma$; and
    \item[(c)] $\forall x\in\X,\exists u\in\U\!:\,\E_{\Tr}[\,\B(X^+)\mid X=x,\,U=u\,] -\B(x) \leq  c;$
  \end{itemize}
  for some constants $\gamma>\eta\geq0$ and $c\geq 0$.
\end{definition}
If one can find a CBC for a system $\M$, then, a lower bound on the probability of $\M$ being safe is given by the following proposition due to \cite{kushner1967stochastic}.
\begin{proposition}[Finite-horizon safety]
  \label{prop:reach}
  Consider an MDP $\M=(\X,\X_0,\U,\Tr)$ and a safety specification $\psi_{\text{safe}}=(\X_u,T)$. Suppose there exists a CBC $B$ w.r.t. $\X_u$ (Definition~\autoref{def:cbc}) with constants $\eta$, $\gamma$, and $c$. Then, there exists a stationary policy $\pi$ such that
  $$P^\pi_{\text{safe}}(\M)\geq 1- \frac{\eta + cT}{\gamma}.$$
\end{proposition}

\paragraph{Implementation}

When verifying the safety of a system, \lucid will search for a CBC $\B$ that satisfies the conditions in Definition~\ref{def:cbc}.
If it finds it, it will return the constants $\eta$, $\gamma$, and $c$, as well as the CBC $\B$ itself, which becomes a certificate of safety for the system $\M$.
The template of the barrier function is explored later in section~\ref{sec:benchmarks}.

\subsection{Kernel Mean Embeddings}

A symmetric function $k_\X:\X\times\X\rightarrow\R$ is called a (positive definite) \emph{kernel} (note the distinction from \emph{probability kernels}) if for all $N\in\N_{>0}$ we have $\sum_{i=1}^{N}\sum_{j=1}^{N}a_i a_j\allowbreak k_\X(x_i,x_j) \geq 0$ for $x_1,\ldots,x_N\in\X\subset{\R^n}$ and $a_1,\ldots,a_N\in\R$.
A prominent example is the \emph{squared exponential} (SQExp) kernel \cite{Rasmussen2005GP,Kanagawa2018GPvsKernel}:
\begin{equation}
  \label{eq:gaussian-kernel}
  k_\X(x,x') := \sigma_f^2 \exp\left( -\frac{1}{2} (x-x')\T \Sigma\, (x-x') \right),\quad \Sigma:=\diag(\sigma_l)^{-2},
\end{equation}
with amplitude $\sigma_f^2\geq0$ and lengthscale coefficients $\sigma_l\in\R^n$.
%In the following, we refer to a positive definite kernel as \emph{kernel}, for simplicity.
For this work, we assume that all kernels are bounded on their domain, i.e., $\E_{}[k_\X(x,x)]<\infty$, $x\in\X$.
Given a kernel $k_\X$ on a non-empty set $\X$, there exists a unique corresponding \emph{reproducing kernel Hilbert space} (RKHS) $\Hilbert_{k_\X}$ of functions $f:\X\rightarrow\R$ equipped with an inner product $\innerH{\cdotx}{\cdotx}{\Hilbert_{k_\X}}$ possessing the \emph{reproducing property} such that for any function $f\in\Hilbert_{k_\X}$ and $x\in\X$ we have $f(x)=\innerH{f}{k_\X(\cdotx,x)}{\Hilbert_{k_\X}}$.
Note that $k_\X(\cdotx,x):\X\rightarrow\Hilbert_{k_\X}$ is a real-valued function,
\new{which is} also called an implicit \emph{canonical} \emph{embedding} or \emph{feature map} $\phi_\X$
such that $k_\X(x,x')=\innerH{\phi_\X(x)}{\phi_\X(x')}{\Hilbert_{k_\X}}$ for all $x,x'\in\X$.
For an RKHS $\Hilbert_{k_\X}$, we use the associated feature map $\phi_\X$ and kernel $k_\X$ interchangeably for ease of notation and comprehensibility.
%We will call the kernel associated to an RKHS $\Hilbert_{k_\X}$ the \emph{characteristic} kernel of $\Hilbert_{k_\X}$ and use $\phi_\X$ and $k_\X$ interchangeably for ease of notation and comprehensibility.
%and an implicit \emph{canonical} \emph{embedding} or \emph{feature map} $\phi_\X:\X\rightarrow\Hilbert_{k_\X}$ with an inner product on $\Hilbert_{k_\X}$, i.e., $k_\X(x,x')=\innerH{\phi_\X(x)}{\phi_\X(x')}{\Hilbert_{k_\X}}$, 
%Furthermore, 
The inner product induces the norm $\norm{f}_{\Hilbert_{k_\X}}\!\!\!\!:=\!\!\sqrt{\smash[b]{\innerH{f}{f}{\Hilbert_{k_\X}}}}$ of the RKHS.
%Note that $\Hilbert_{k_\X}$ is a space of functions $f:\X\rightarrow\R$. 
%It is equipped with the infamous \emph{reproducing property} such that for any function $f\in\Hilbert_{k_\X}$ and $x\in\X$ we have $f(x)=\innerH{f}{\phi_\X(x)}{\Hilbert_{k_\X}}$.
%Furthermore, the norm associated with the RKHS $\Hilbert_{k_\X}$ is defined as $\norm{f}_{\Hilbert_{k_\X}}:=\sqrt{\innerH{f}{f}{\Hilbert_{k_\X}}}$.
%We refer the interested reader to XX for a comprehensive dissemination on RKHS.
Throughout this paper, we assume that all RKHSs are \emph{separable}.
Given $N$ i.i.d. samples $\hat X_N:=[\hat{x}_i]_{i=1}^N$ with $\hat{x}_i\in\X$, the
\emph{Gram matrix} of $k_\X$ is given by
$K_{\hat{X}}^N:=[k_\X(\hat{x}_i,\hat{x}_j)]_{i,j=1}^N.$
Furthermore, we define the vector-valued function
$k_{\hat{X}}^N(x)  := [k_\X(x,\hat{x}_i)]_{i=1}^N.$

\subsection{Data-Driven Barrier Formulation}
\label{sec:data-driven-barrier-formulation}
Compared to the typical spatial form \eqref{eq:representer-form} based on spatial support vectors $\bar{x}_1,\bar{x}_2,\ldots\in\X$, the spectral form \eqref{eq:representer-form-spectral} is a much more efficient representation for higher-dimensional systems.
The number of spatial support vectors needed to construct adequately expressive barriers grows exponentially in the dimensionality of the system. We avoid this.

Apart from the memory efficiency of the spectral barrier form, we can apply the same idea to collapse the complex CME term in \eqref{eq:semiinf-prog-kushner} via the \emph{fast Fourier transform} (FFT). That is we obtain spectral coefficients $H\in\R^{(2M+1)\times (2M+1)}$ s.t.
\begin{equation}
  k^N_{\hat{X}}(x)\T\left[ K^N_{\hat{X}}+N\lambda I_N\right]^{-1\!} (\Phi_{M, \hat{X}^{+\!}}^{N})\T b \approx \phi_M(x)\T H b,\label{eq:cme-term-spectral-approx}
\end{equation}
where $\Phi_{M, \hat{X}^{+\!}}^{N} := [\phi_M(\hat{x}^+_1),\ldots,\phi_M(\hat{x}^+_N)]$.
Whilst projection onto a truncated Fourier basis introduces an approximation error, the error decreases exponentially with the cardinality of wavenumbers in the basis \cite{rahimi2007random}.
% Our numerical results suggest that the empirical error is negligible.

>>>>

%File: anonymous-submission-latex-2025.tex
\documentclass[letterpaper]{article} % DO NOT CHANGE THIS

@@

% Title

% Your title must be in mixed case, not sentence case.
% That means all verbs (including short verbs like be, is, using,and go),
% nouns, adverbs, adjectives should be capitalized, including both words in hyphenated terms, while
% articles, conjunctions, and prepositions are lower case unless they
% directly follow a colon or long dash
\title{LUCID: Learning-Enabled Uncertainty-Aware Certification of Stochastic Dynamical Systems}
%\title{LUCID -- ???}
% Black-Box Verification Engine for Embodied AI Systems?
% \author {
%     % Authors
    % Ernesto Casablanca\textsuperscript{\rm 1},
    % Oliver Sch\"on\textsuperscript{\rm 1}%,
    % Third Author Name\textsuperscript{\rm 2}
% }
% \affiliations {
    % Affiliations
    % \textsuperscript{\rm 1}Newcastle University, United Kingdom\\
    % \textsuperscript{\rm 2}Affiliation 2\\
    % e.casablanca2@ncl.ac.uk, o.schoen2@ncl.ac.uk%, thirdAuthor@affiliation1.com
% }

% REMOVE THIS: bibentry
% This is only needed to show inline citations in the guidelines document. You should not need it and can safely delete it.
\usepackage{bibentry}
% END REMOVE bibentry

@@

@@

\begin{abstract}
    Ensuring the safety of AI-enabled systems, particularly in high-stakes domains such as autonomous driving and healthcare, has become increasingly critical.
    Traditional formal verification tools fall short when faced with systems that embed both opaque, black-box AI components and complex stochastic dynamics.
    To address these challenges, we introduce \lucid (Learning-enabled Uncertainty-aware Certification of stochastIc Dynamical systems), a verification engine for certifying safety of black-box stochastic dynamical systems from a finite dataset of random state transitions.
    As such, \lucid is the first known tool capable of establishing quantified safety guarantees for such systems.
    Thanks to its modular architecture and extensive documentation, \lucid is designed for easy extensibility.

    \lucid employs a data-driven methodology rooted in control barrier certificates, which are learned directly from system transition data, to ensure formal safety guarantees.
    We use conditional mean embeddings to embed data into a reproducing kernel Hilbert space (RKHS), where an RKHS ambiguity set is constructed that can be inflated to robustify the result to out-of-distribution behavior.

    A key innovation within \lucid is its use of a finite Fourier kernel expansion to reformulate a semi-infinite non-convex optimization problem into a tractable linear program.
    The resulting spectral barrier allows us to leverage the fast Fourier transform to generate the relaxed problem efficiently, offering a scalable yet distributionally robust framework for verifying safety.
    \lucid thus offers a robust and efficient verification framework, able to handle the complexities of modern black-box systems while providing formal guarantees of safety.
    These unique capabilities are demonstrated on challenging benchmarks.
\end{abstract}

% Uncomment the following to link to your code, datasets, an extended version or similar.
%
\begin{links}
    \link{Source Code}{https://gitlab.com/lucidtoolsource/lucid}
    \link{Documentation}{https://lucidtoolsource.gitlab.io/lucid/}
    % \link{Datasets}{https://aaai.org/example/datasets}
    % \link{Extended version}{https://aaai.org/example/extended-version}
\end{links}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Introduction}
Embodied forms of AI are on the rise, including applications such as autonomous vehicles, robotics, personalized healthcare, and smart infrastructure.
Their core functionality is build around the advances of deep learning, enabling systems to understand and reason in complex and human-like ways to produce a desired behavior.
Whilst the black-box nature of AI components has been a crucial contributor to the widespread success of deep learning, in response to a rapidly evolving legal scrutinization~\cite{veale2021demystifying}, the resulting lack of traceability and certifiability has put a premature halt to its deployment in safety-critical applications.

As deep learning agents are left to choose their actions autonomously in closed-loop interaction with the physical world, they are confronted with a world riddled with randomness and uncertainty. 
Efforts to establishing trust in their safe operation have been largely focused on developing tools to verify the input--output behavior of neural networks (NNs) embedded in the systems~\cite{liu2021algorithms}.
Unfortunately, there do not exist any tools that could take these results and certify the safety of the entire system, as uncertainty-aware models of closed-loop systems are rarely available and simulators are often too complicated and opaque to perform verification directly~\cite{wongpiromsarn2023formal}. This motivates a holistic black-box treatment~\cite{corso2021survey}, where safety guarantees are to be established from behavioral data and in account of the unknown uncertainty.

There exist no tools capable of solving this problem with quantified guarantees.
%
For the setting where a model of the closed-loop system is available, \npinterval by \citet{harapanahalli2023forward} verifies safety of nonlinear systems with non-deterministic disturbance based on existing NN verification tools. See the references therein and the annual friendly competition by \citet{abate2024arch} for adjacent work.
%
Few data-driven approaches for stochastic systems exist. 
\omnisafe is a comprehensive platform for the development of safe reinforcement learning (RL) algorithms~\citet{ji2024omnisafe}. However, safe RL does generally only \emph{encourage} safer behavior, without providing any rigorous or quantified guarantees on the probability of the absence of unsafe behavior.
As a popular working principle for certifying safety, control barrier certificates (CBCs)~\cite{prajna2006barrier} are at the basis of tools such as \trust~\cite{gardner2025trust} (supports only polynomial dynamics) and \fossil~\cite{edwards2024fossil} (model-based), which are restricted to deterministic systems.

To close this gap (see Table~\ref{tab:qualitative_comparison}), we introduce \lucid, the first verification engine for black-box stochastic dynamical systems with quantified guarantees.
\begin{table}
    \centering
    \begin{tabular}{ccccccc}
        \toprule
        \textbf{Tool}                                    & \multicolumn{6}{c}{\textbf{Supported Features}}                                                                                                                                                             \\[.4em]
                                                         & \rotatebox{90}{Guarantees}                      & \rotatebox{90}{Data Driven} & \rotatebox{90}{Stochastic D.} & \rotatebox{90}{Non-Poly. D.} & \rotatebox{90}{Stat. Correct.} & \rotatebox{90}{Closed Loop} \\
        \midrule
        \textbf{\lucid}                                  & \cmark                                          & \cmark                      & \cmark                        & \cmark                       & \cmark                         & \cmark                      \\
        \trust~\shortcite{gardner2025trust}              & \cmark                                          & \cmark                      & \xmarkred                     & \xmarkred                    & \cmark\footnotemark[1]            & \cmark                      \\
        \fossil~\shortcite{edwards2024fossil}            & \cmark                                          & \xmarkred                   & \xmarkred                     & \cmark                       & NA                             & \cmark                      \\
        \omnisafe~\shortcite{ji2024omnisafe}             & \xmarkred                                       & \cmark                      & \cmark                        & \cmark                       & \xmarkred                      & \cmark                      \\
        \npinterval~\shortcite{harapanahalli2023forward} & \cmark                                          & \xmarkred                   & \cmark/\xmarkred\footnotemark[2] & \cmark                            & NA                             & \cmark                      \\
        % \deepsplit~\shortcite{henriksen2021deepsplit} & \xmarkred & \xmarkred & \cmark & \xmarkred& \xmarkred \\
        % \reluplex~\shortcite{katz2017reluplex} & \xmarkred & \xmarkred & \cmark & \xmarkred & \xmarkred \\
        % tool6                                            &                                                 &                             &                               &                                                                                             \\
        \bottomrule
    \end{tabular}
    \caption{Qualitative comparison with existing tools based on their supported features: quantified safety guarantees, data driven, stochastic dynamics, non-polynomial dynamics, statistical correctness guarantees of the learned model (only applicable if data driven), and support for closed-loop systems.}
    \label{tab:qualitative_comparison}
\end{table}
\footnotetext[1]{Assuming the data satisfies persistence of excitation.} % TRUST
\footnotetext[2]{Accepts non-deterministic bounded disturbances.} % npinterval
At its core, it learns CBCs for unknown systems based solely on data, by constructing an uncertainty-aware estimator of the expected system behavior based on \emph{conditional mean embeddings} (CMEs)~\cite{muandet2017kernel}.
Crucially, the underlying learning framework establishes quantified safety probabilities with distributionally robust guarantees for arbitrary smooth dynamics, thus relaxing the need for restrictive structural assumptions common to related approaches.
For instance, \citet{schoen2024distributionally} learn CBCs for systems with polynomial dynamics and \citet{chen2025distributionally} assume systems with known deterministic component.
%
Alternative approaches often leave the statistical correctness with respect to the underlying data-generating process unaddressed: \citet{kazemi2020fullLTL} use model-free reinforcement learning, relying on known Lipschitz constants; \citet{lew2021sampling} compute reachable sets for stochastic systems using a sampling-based scheme, which yields only asymptotic guarantees; and \citet{salamati2024data} use a scenario-based method, based on Lipschitz constants and exponentially large datasets.
For data-driven safety verification of stochastic systems via conformal prediction~\cite{lindemann2024formal}, no tools are available.

We summarize the main contributions of this work:
\begin{itemize}
    \item We introduce \lucid, a novel verification engine that learns control barrier certificates from data using kernel-based CMEs. \lucid uses a tractable reformulation of the verification problem via a finite Fourier expansion, enabling efficient barrier synthesis based on a linear program.
    \item A robust and extensible software implementation, with both C++ and Python interfaces, supporting configuration via YAML, JSON, or Python scripts, and offering both a \gls{cli} and a \gls{gui}.
    \item Numerical evaluation on a suite of benchmarks, demonstrating \lucid{}’s unique capabilities, robustness, and practical utility.
\end{itemize}

\paragraph{Organization}
The paper is organized as follows:
Section~\ref{sec:theory} provides the theoretical foundations of \lucid, including the safety of black-box dynamical systems, control barrier certificates, conditional mean embeddings, and derivation of the relaxed linear program.
Section~\ref{sec:tool} describes the architecture and functionalities of \lucid alongside a running example, presenting \lucid's components and how they interact.
% Section~\ref{sec:example} presents a simple example use case, demonstrating how to use \lucid to certify the safety of a simple linear system.
Section~\ref{sec:experiments} evaluates \lucid on a suite of benchmarks. Finally, Sections~\ref{sec:future_extensions} and \ref{sec:conclusion} provide future extensions and a conclusion.

\section{Theoretical Working Principles}\label{sec:theory}
\subsection{Safety of Black-Box Dynamical Systems}

\paragraph{System Description}
Many AI-driven systems exhibit complex, nonlinear, and stochastic behavior that can be modeled as discrete-time stochastic processes with Markovian dynamics:
\begin{equation}
    \label{eq:model}
    \S\colon\quad x_{t+1}  = f(x_t,a_t,w_t),\quad w_t\sim p_w,
\end{equation}
where $f\colon\X\times\A\times\W\rightarrow\X$ is a continuous vector field describing the evolution of the system state $x_t\in\X\subset\R^n$ over time $t\in\N_{\geq0}$, driven by control actions $a_t\in\A\subset\R^m$ and process noise $w_t\in\W\subset\R^l$.
The noise is assumed to be drawn from a stochastic distribution $p_w$ in an independent and identically distributed (i.i.d.) manner.
%
This general formulation subsumes a wide range of systems, including discrete-time Markov decision processes (MDPs).

\paragraph{Black-Box Policies}
Here, the focus is on systems $\S$ driven by black-box control policies of the form $\pi\colon\R^n\rightarrow\R^m$, i.e., at every time step $t=0,1,2,\ldots$ a continuous action $a_t\in\A$ is selected based on the current state $x_t$. Such policies could be, for example, neural networks trained via RL, or any other black-box function generating actions in $\R^m$ (see Figure~\ref{fig:closed_loop_system}).
The resulting closed-loop system is denoted as $\S^\pi$.
\begin{figure}
    \centering
    \begin{tikzpicture}
        \foreach \i in {4, ..., 1}{
                \node[name=xa\i] at (\i/4*9,0) {$x_\i$};
            }
        \foreach \i in {4, ..., 1}{
                \node[name=aa\i] at (\i/4*9+.5,0) {$u_{\i}$};
            }
        \foreach \i in {4, ..., 1}{
                \path[draw,->] (xa\i) -- +(0,.5)-| node[left,above,xshift = -0.3cm]{\footnotesize$\pi$} (aa\i);
            }
        \foreach \i in {3, ..., 1}{
                \pgfmathtruncatemacro{\j}{\i + 1}
                \path[draw,->] (aa\i)-- node[above]{\footnotesize$\S$} (xa\j);
            }
    \end{tikzpicture}
    \caption{Evolution of the closed-loop system.}
    \label{fig:closed_loop_system}
\end{figure}

\paragraph{Dataset}
Due to their complexity and opacity, such systems must often be treated as black boxes in their entirety, assuming only access to a finite amount, $N$, of system observations
\begin{equation}
    \data_N\colon \quad \{x^i,a^i,x^i_+\}_{i=1}^N,\label{eq:data}
\end{equation}
where every sample transition from $x^i\in\X$ to a successor $x^i_+\in\X$ is generated as a realization
\begin{equation*}
    (x^i,a^i,x_+^i)\sim\int \delta_{f(x^i,a^i,w)}\,p_w(dw)\,\mathcal{U}_\X(dx^i)\,\mathcal{U}_\A(da^i),
\end{equation*}
with $\mathcal{U}_\X,\mathcal{U}_\A$ uniform distributions on $\X$ and $\A$, respectively.

\input{figures/steps}

\paragraph{Safety Problem}
A question that often arises when designing a policy $\pi$ for a system $\S$, especially in engineering contexts, is if the closed-loop system $\S^\pi$ elicits safe behavior. That is, when operating in some operating domain $\X_{0}\subset\X$ under a policy $\pi$, the system $\S^\pi$ shall avoid unsafe regions $\X_U\subset\X$ (such as obstacles) for at least some predefined control horizon $T\in\N\cup\{\infty\}$.

Such safety problems are tricky and even for the case where an explicit model \eqref{eq:model} is given, certifying safety of a discrete-time stochastic system over an uncountable state space does generally not admit an analytical solution and is extremely challenging, especially for complex dynamics~\cite{abate2008probabilistic}.
Even more, for unbounded stochasticity $w_t\sim p_w$, there is generally no deterministic answer to safe or not, but a safety probability $P_{\text{safe}}(\S^\pi)\in[0,1)$. Thus the goal is generally to estimate a \emph{lower bound} on the true safety probability $P_{\text{safe}}(\S^\pi)$.

\paragraph{Problem Statement}
Assuming access to a finite dataset $\data_N$ from the black-box system $\S$ in \eqref{eq:model}, quantify a certifiable lower bound on the probability of the black-box system $\S^\pi$ being safe with respect to a safety specification given by $(\X_0,\X_U,T)$.

\medskip

\lucid provides a solution to this problem by automating the steps in Figure~\ref{fig:steps}, outlined in the next sections.

\subsection{Control Barrier Certificates}\label{sec:cbc}
\glspl{cbc}\footnote[3]{\glspl{cbc} and discrete-time \glspl{cbf} \cite{cosner2023generative} are equivalent.} leverage the concept of set invariance to arrive at an abstraction-free numerical solution to finding a lower bound on $P_{\text{safe}}(\S^\pi)$. This has made them popular tools for safety verification and controller synthesis~\cite{prajna2006barrier}.

A non-negative function $\B\colon\X\rightarrow\R_{\geq 0}$ is a \gls{cbc} of a system $\S$ with reference to an unsafe set $\X_U$, if it satisfies
\begin{itemize}
    \item[(a)] $\forall x_0\in \X_0\colon\,\B(x_0)\leq\eta$;
    \item[(b)] $\forall x_U\in \X_U\colon\,\B(x_u)\geq\gamma$; and
    \item[(c)] $\forall x\in\X,\,\exists a\in\A\colon\,\E[\B(X_+) \mid X=x,\,A=a] -\B(x) \leq  c;$
\end{itemize}
for some constants $\gamma>\eta\geq0$ and $c\geq 0$. Here, upper case $X_+$, $X$, and $A$ denote the random variables underlying concrete realizations $(x_t,a_t,x_{t+1})$ elicited by $\S$.
%
Intuitively, if one can find a \gls{cbc} for a system $\S$, then, a lower bound on the probability of $\S$ being safe can be quantified based on the distance between the two level sets $\gamma$ and $\eta$~\cite{kushner1967stochastic}:
\begin{equation}
    P_{\text{safe}}(\S^\pi)\geq 1- \frac{\eta + cT}{\gamma},\label{eq:safety_prob_lb}
\end{equation}
where $T$ is the desired control horizon.
Notably, the system dynamics only enter the barrier conditions through the conditional expectation in (c), and a \gls{cbc} thus only depends on the \emph{expected} system behavior, with no need to consider its concrete stochastic law.

Whilst \eqref{eq:safety_prob_lb} can provide a robust assessment of a systems safety, in practice, the bound can be overly conservative and thus several improved variants of the original barrier constraints exist~\cite{mahathi2022kinductive}.
As these conditions in essence all rely on the computation of chance constraint with respect to the expected behavior of the system, they are basically interchangeable.

Although there exist model-\emph{based} efficient solutions to this problem for linear and control affine systems, especially for complex systems this is generally a semi-infinite problem, demanding a data-driven solution. Furthermore, for the data-driven case, establishing constraint (c) rigorously without relying on impractical assumptions is extremely challenging.
